---
title: "Multinational Time Use Study Analysis"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

# Multinational Time Use Study Analysis of Active Transport

This repository contains just one script used to analyse changes in transport
behaviour over time. The data were obtained from the [Multinational Time Use
Study (MTUS)](https://www.ipums.org/timeuse.shtml), part of the [Integrated
Public Use Microdata Series (IPUMS)](https://www.ipums.org). MTUS data are
jointly hosted by the Minnesota Population Center at the University of
Minnesota, and the [Centre for Time Use Research](https://www.timeuse.org/mtus)
at the University of Oxford.

Data downloads are accompanied by a corresponding "ddi" file, included here both
in [`.xml`](./data/mtus_00001.xml) and [plain text](./data/mtus_00001.cbk)
formats, as well as an equivalent [`.html` document](./data/mtus_00001.dat.html)
for easy reading. This file codifies the hierarchical structure of the data,
including descriptions of all variable encodings.  The data download can be
reproduced by following the structure of these files.

The corresponding data are very large, and therefore not directly included
within this repository. They include the following variables and corresponding
codes (where "MAIN" describes the main activity):

| variable | code | description |
| --- | --- | --- |
| MAIN | 043 | walking |
| MAIN | 044 | cycling |
| MAIN | 063 | travel to/from work |
| MAIN | 064 | education travel |
| MAIN | 065 | voluntary/civic/religious travel |
| MAIN | 066 | child/adult care travel |
| MAIN | 067 | shop, person/child care travel |
| MAIN | 068 | other travel |
| MTRAV | 01 | car |
| MTRAV | 02 | public transport |
| MTRAV | 03 | walk/on foot |
| MTRAV | 04 | other physical transport |
| MTRAV | 05 | other/unspecified |

## Reading the data

IPUMS data can now be read directly into **R** using the fabulous [`ipumsr`
package](https://github.com/mnpopcenter/ipumsr). The primary functions simply
need the "ddi" file describing the data structure. It is presumed here that the
data themselves reside in the same directory (generally in `.gz` form).

```{r read-data}
library (ipumsr)
dat <- read_ipums_micro_list (ddi = "./data/mtus_00001.xml",
                              vars = c ("COUNTRY", "YEAR", "AGE", "SEX",
                                        "TIME", "MAIN", "MTRAV"))
dat
```
Note that the `dat$ACTIVITY$TIME` variable quantifies the duration of a travel
activity, and is the key variable we will be analysing here.  These data include
over 9 million activities from nearly 500,000 individuals from these countries:
```{r}
countries <- unique (dat$PERSON$COUNTRY)
countries
```
The following analyses will use `tidyverse` functionality, so before we start we
load all packages:
```{r tidyverse, message = FALSE}
require (tidyverse)
```

## Filter to active transport data

The two tables represent the hierarchical structure of the data. These data can
then be filtered down to only those records describing transport activities.  We
will only use the `ACTIVITY` table here.
```{r filter-activity}
act <- filter (dat$ACTIVITY, MTRAV > 0 | MAIN %in% c (43:44, 63:68))
act
```
We still have almost 1.5 million travel activities.  Now let's make equivalent
tables for each activity, plus one for all travel activities combined
(represented by `MTRAV > 0`):
```{r filter-activity2}
bike <- filter (act, MAIN == 44 | MTRAV == 4)
walk <- filter (act, MAIN == 43 | MTRAV == 3)
allmodes <- filter (act, MTRAV > 0)
```
Note that `MTRAV == 4` describes "other physical transport," which may or may
not represent bicycle-borne transport.

## Aggregate activity times

The data in the resultant tables then need to be aggregated to generate total
durations of each activity for each year and country. The following function
converts those durations to proportions of cycling and walking as a fraction of
total travel time.
```{r gettimes-fn}
gettimes <- function (ci = "CA")
{
    tb <- bike %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    tw <- walk %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    ta <- allmodes %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    if (nrow (tb) == 0)
        message (ci, ": no bike data")
    if (nrow (tw) == 0)
        message (ci, ": no walking data")
    year <- sort (unique (c (tb$YEAR, tw$YEAR, ta$YEAR)))
    year <- year [which (year %in% ta$YEAR)]

    times <- tibble (year = year,
                     walk = tw$time [match (year, tw$YEAR)],
                     bike = tb$time [match (year, tb$YEAR)],
                     allmodes = ta$time [match (year, ta$YEAR)])
    times [is.na (times)] <- 0
    times$walk <- times$walk / times$allmodes
    times$bike <- times$bike / times$allmodes
    if (nrow (times) < 3) # require >= 3 years of data
        message (ci, ": no multi-year data")
    return (times)
}
times <- lapply (countries, gettimes)
names (times) <- countries
```
Most countries do not unfortunately have sufficient data to analyse further,
leaving only the UK, the US, and the Netherlands.
```{r filter-to-ukus}
times <- times [which (countries %in% c ("UK", "US", "NL"))]
times
```
The US also only has usable data after 1998, and the Netherlands after 1975, so
```{r filter-us-times}
times$US <- filter (times$US, year > 1998)
times$NL <- filter (times$NL, year > 1975)
```

## Final data wrangling and plotting

These data can now be converted to a single `data.frame` to be plotted
```{r bind-countries}
times$UK$COUNTRY <- "UK"
times$US$COUNTRY <- "US"
times$NL$COUNTRY <- "NL"
times <- do.call (rbind, times)
```
Bicycle journeys in the US and UK generally represent around one fifth of the
proportion of walking journeys, so we multiply them by 5 to enable both to be
plotted on the same scale
```{r bike-times-5}
times$bike [times$COUNTRY != "NL"] <- times$bike [times$COUNTRY != "NL"] * 5
```
Now merge the variables into a single `country_trans_mode` variable for easy
plotting:
```{r gather, message = FALSE}
times <- gather (times, key = trans_mode, value = time, walk, bike)
times$country_mode <- paste0 (times$COUNTRY, ":", times$trans_mode)
```
These data may now be used to plot changes in rates of walking and cycling over
time, using the [`solarized` colours](http://ethanschoonover.com/solarized) from
the [`ggthemes` package](https://github.com/jrnold/ggthemes).
```{r plot, message = FALSE}
require (ggthemes)
ggplot (times, aes (year, time, colour = country_mode)) +
    geom_line () +
    geom_point () +
    theme_solarized ()
```
Remember that rates of bicycle trips are multiplied by 5 for the UK and US, but
not for the NL. The drastic decreases in the final points for the UK (2005) seem
anomalous, because it is unlikely that walking actually decreased by almost one
half. If we consider all other results to be reliable, they suggest that
patterns in the Netherlands have remained largely stable over 20 years, with a
slight decrease in rates of walking, and a notable increase in cycling between
2000 and 2005. Active transport generally decreased in the US until 2000, after
which both walking and cycling have progressively increased to respectively
represent around 7 and just over 1%. Rates in the UK appear, in contrast, to
have increased much more rapidly.

The following code quantifies the relative rates of change in the US post-2000,
and in the UK up to and including 2001. The analysis will be grouped by
`COUNTRY` and `trans_mode`, requiring the following lines to prepare the data
for analysis:
```{r filter-times}
timesf <- filter (times, !((country_mode == "UK:walk" & year == 2005) |
                           (country_mode == "UK:bike" & year == 2005) |
                           (country_mode == "US:walk" & year <= 2000) |
                           (country_mode == "US:bike" & year <= 2000))) %>%
                group_by (COUNTRY, trans_mode) %>%
                nest ()
```
The analysis can then be implemented with the following code:
```{r model-times}
mod <- function (df)
    summary (lm (time ~ year, data = df))
timesf <- timesf %>%
    mutate (mod = map (timesf$data, mod)) %>%
    mutate (coeffs = mod %>% map (broom::tidy)) %>%
    unnest (coeffs) %>%
    filter (term == "year")
```
The second `mutate` call maps the `lm` models to `data.frame` objects of
coefficients, which generates a `term` column containing values for both
`(Intercept)`, and `year`, where the last line filters only the latter. Finally,
converts rates of change to annual percentages, and remove a couple of
statistics to help us concentrate on what we want to know. Rates of change for
cycling are also divided by 5 to account for the previous multiplication that
was used for plotting.
```{r}
timesf$std.error <- timesf$statistic <- NULL
timesf$estimate <- timesf$estimate * 100
timesf$estimate [timesf$COUNTRY != "NL" & timesf$trans_mode == "bike"] <- 
    timesf$estimate [timesf$COUNTRY != "NL" & timesf$trans_mode == "bike"] / 5
timesf
```
All changes are significant except for those in the Netherlands. Rates of
walking increased by almost 1% per year in the UK, and just over 0.1% in the US.
In contrast to these changes in rates of walking, rates of cycling increased
relatively much more rapidly in the US (0.04% per year) than in the UK (0.11%
per year).
