---
title: "Multinational Time Use Study Analysis"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

# Multinational Time Use Study Analysis of Active Transport

This repository contains just one script used to analyse changes in transport
behaviour of time. The data were obtained from the [Multinational Time Use Study
(MTUS)](https://www.ipums.org/timeuse.shtml) data, part of the [Integrated
Public Use Microdata Series (IPUMS)](https://www.ipums.org). MTUS data are
jointly hosted by the Minnesota Population Center at the University of
Minnesota, and the [Centre for Time Use Research](https://www.timeuse.org/mtus)
at the University of Oxford.

Data downloads are accompanied by a corresponding "ddi" file, included here both
in [`.xml`](./data/mtus_00001.xml) and [plain text](./data/mtus_00001.cbk)
formats. The data download can be reproduced by following the structure of these
files.

The downloaded data include the following variables and corresponding codes
(where "MAIN" describes the main activity):

| variable | code | description |
| --- | --- | --- |
| MAIN | 043 | walking |
| MAIN | 044 | cycling |
| MAIN | 063 | travel to/from work |
| MAIN | 064 | education travel |
| MAIN | 065 | voluntary/civic/religious travel |
| MAIN | 066 | child/adult care travel |
| MAIN | 067 | shop, person/child care travel |
| MAIN | 068 | other travel |
| MTRAV | 01 | car |
| MTRAV | 02 | public transport |
| MTRAV | 03 | walk/on foot |
| MTRAV | 04 | other physical transport |
| MTRAV | 05 | other/unspecified |

## Reading the data

IPUMS data can now be read directly into **R** using the fabulous [`ipumsr`
package](https://github.com/mnpopcenter/ipumsr). The primary functions simply
need the "ddi" file describing the data structure. It is presumed here that the
data themselves reside in the same directory. (They are not included in this
repository because of their very large size).

```{r read-data}
library (ipumsr)
dat <- read_ipums_micro_list (ddi = "./data/mtus_00001.xml",
                              vars = c ("COUNTRY", "YEAR", "AGE", "SEX",
                                        "TIME", "MAIN", "MTRAV"))
dat
```
Note that the `dat$ACTIVITY$TIME` variable quantifies the duration of a travel
activity, and is the key variable we will be analysing here.  These data include
over 9 million activities from nearly 500,000 individuals from these countries:
```{r}
countries <- unique (dat$PERSON$COUNTRY)
countries
```

## Filter to active transport data

The two tables represent the hierarchical structure of the data. These data can
then be filtered down to only those records in which people walked, rode
bicycles, or used cars. We will only use the `ACTIVITY` table here.
```{r filter-activity}
act <- dplyr::filter (dat$ACTIVITY, MTRAV > 0 | MAIN %in% c (43:44, 63:68))
act
```
We still have almost 1.5 million travel activities.  Now let's make equivalent
tables for each activity, plus one for all travel activities combined
(represented by `MTRAV > 0`):
```{r filter-activity2}
bike <- dplyr::filter (act, MAIN == 44)
walk <- dplyr::filter (act, MAIN == 43)
allmodes <- dplyr::filter (act, MTRAV > 0)
```

## Aggregate activity times

Then aggregate those data to generate total durations of each activity for each
year and country. The following function converts those durations to
proportional times, enabling proportions of cycling and walking to be directly
compared in relation to total travel time. (The latter is admittedly a rough
estimate
```{r gettimes-fn}
require (magrittr)
require (tibble)
gettimes <- function (ci = "CA")
{
    tb <- bike %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    tw <- walk %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    ta <- allmodes %>%
        dplyr::filter (COUNTRY == ci) %>%
        dplyr::group_by (YEAR) %>%
        dplyr::summarize (time = sum (TIME))
    if (nrow (tb) == 0)
        message (ci, ": no bike data")
    if (nrow (tw) == 0)
        message (ci, ": no walking data")
    year <- sort (unique (c (tb$YEAR, tw$YEAR, ta$YEAR)))
    if (length (year) < 2)
        message (ci, ": no multi-year data")
    year <- year [which (year %in% ta$YEAR)]

    times <- tibble (year = year,
                     walk = tw$time [match (year, tw$YEAR)],
                     bike = tb$time [match (year, tb$YEAR)],
                     allmodes = ta$time [match (year, ta$YEAR)])
    times [is.na (times)] <- 0
    times$walk <- times$walk / times$allmodes
    times$bike <- times$bike / times$allmodes
    return (times)
}
times <- lapply (countries, gettimes)
names (times) <- countries
```
Most countries do not unfortunately have sufficient data to analyse further,
leaving only the UK and the US.
```{r filter-to-ukus}
times <- times [which (countries %in% c ("UK", "US"))]
times
```
The US also only has usable data after 1998, so
```{r filter-us-times}
times$US <- dplyr::filter (times$US, year > 1998)
```

## Final data wrangling and plotting

These data can now be converted to a single `data.frame` to be plotted
```{r bind-countries}
times$UK$COUNTRY <- "UK"
times$US$COUNTRY <- "US"
times <- do.call (rbind, times)
```
Bicycle journeys generally represent around one fifth of the proportion of
walking journeys, so we multiply them by 5 to enable both to be plotted on the
same scale
```{r bike-times-5}
times$bike <- times$bike * 5
```
Now merge the variables into a single `country_trans_mode` variable for easy
plotting:
```{r gather, message = FALSE}
require (tidyr)
times <- gather (times, key = trans_mode, value = time, walk, bike)
times$country_mode <- paste0 (times$COUNTRY, ":", times$trans_mode)
```
These data may now be used to plot changes in rates of walking and cycling over
time, using the [`solarized` colours](http://ethanschoonover.com/solarized) from
the [`ggthemes` package](https://github.com/jrnold/ggthemes).
```{r plot, message = FALSE}
require (ggplot2)
require (ggthemes)
ggplot (times, aes (year, time, colour = country_mode)) +
    geom_line () +
    geom_point () +
    theme_solarized ()
```
